---
title: "Distributional Regression with _bamlss_"
author: "Nikolaus Umlauf, Nadja Klein, Achim Zeileis, Thorsten Simon"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{First Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{distributional regression, first steps}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE, results="hide"}
library("bamlss")
data("mcycle", package = "MASS")
f <- list(accel ~ s(times, k = 20), sigma ~ s(times, k = 20))
if(!file.exists("toymodel.rda")) {
  set.seed(123)
  b <- bamlss(f, data = mcycle, family = "gaussian",
    n.iter = 12000, burnin = 2000, thin = 10)
} else {
  load("toymodel.rda")
}
prefix <- "http://bayesr.R-Forge.R-project.org/articles/" ## ""
```

## Overview

The R package _bamlss_ provides a modular computational framework for distributional regression
models (and beyond). The implementation follows the conceptional framework presented in
@bamlss:Umlauf+Klein+Zeileis:2017, which supports Bayesian and/or frequentist estimation engines
using complex possibly nonlinear model terms of any type. The highlights of the package are:  

* A unified model description where a `formula` specifies how to set up the predictors
  from the `data` and `family` objects provide information about the response distribution.
* A generic method for setting up model terms and a `model.frame` along with the
  corresponding prior structures. A `transformer` can optionally set up modified terms, e.g.,
  using mixed model representation for smooth terms.
* Support for modular and exchangeable updating functions or complete model fitting engines
  in order to optionally implement either algorithms for maximization of the log-posterior for
  posterior mode estimation or for solving high-dimensional integrals, e.g., for posterior mean
  or median estimation.
  First, an (optional) `optimizer` function can be run, e.g., for computing posterior mode
  estimates. Second, a `sampler` is employed for full Bayesian inference with MCMC, which uses the
  posterior mode estimates from the `optimizer` as staring values. An additional step can be used
  for preparing the `results`.
* Standard post-modeling extractor functions to create sampling statistics, visualizations,
  predictions, etc.

More detailed overviews and examples are provided in the articles:

* [Available Model Terms](`r prefix`terms.html)
* [Generalized Linear Models (+)](`r prefix`glm.html)
* [BAMLSS Families](`r prefix`families.html)
* [Visualization with distreg.vis](`r prefix`distregvis.html)
* [Munich Rent Model](`r prefix`rent.html)
* [Bivariate Gaussian models for wind vectors](`r prefix`bivnorm.html)
* [LASSO-Type Penalization in the Framework of Generalized Additive Models for Location, Scale and Shape](`r prefix`lasso.html)
* [Estimation Engines](`r prefix`engines.html)
* [BAMLSS Model Frame](`r prefix`bf.html)

## Installation

The stable release version of _bamlss_ is hosted on the Comprehensive R Archive Network
(CRAN) at <https://CRAN.R-project.org/package=bamlss> and can be installed via

```{r installation-cran, eval=FALSE}
install.packages("bamlss")
```

The development version of _bamlss_ is hosted on R-Forge at
<https://R-Forge.R-project.org/projects/bayesr/> in a Subversion (SVN) repository.
It can be installed via

```{r installation-rforge, eval=FALSE}
install.packages("bamlss", repos = "http://R-Forge.R-project.org")
```

## Running your first model

As a startup on how to use _bamlss_, we illustrate the basic steps on a small textbook
example using the well-known simulated motorcycle accident data [@bamlss:Silverman:1985].
The data contain measurements of the head acceleration
(in $g$, variable `accel`) in a simulated motorcycle accident, recorded in milliseconds after
impact (variable `times`).
```{r}
data("mcycle", package = "MASS")
print(head(mcycle))
```
To estimate a Gaussian location-scale model with
$$
\texttt{accel} \sim \mathcal{N}(\mu = f(\texttt{times}), \log(\sigma) = f(\texttt{times}))
$$
we use the following model formula
```{r}
f <- list(accel ~ s(times, k = 20), sigma ~ s(times, k = 20))
```
where `s()` is the smooth term constructor from the _mgcv_ [@bamlss:Wood:2018]. Note,
that formulae are provided as `list`s of formulae, i.e., each list entry represents one
parameter of the response distribution. Also note that all
smooth terms, i.e., `te()`, `ti()`, etc., are supported by _bamlss_. This way, it is also
possible to incorporate user defined model terms. A full Bayesian model is the
estimated with
```{r, eval=FALSE}
library("bamlss")

set.seed(123)

b <- bamlss(f, data = mcycle, family = "gaussian",
  n.iter = 12000, burnin = 2000, thin = 10)
```
using `12000` iterations for the MCMC chain, a burnin of `2000` (dropped samples) and a thinning
of `10`, i.e., only every 10th sample is saved. Note that per defaul `bamlss()` uses a backfitting
algorithm to compute posterior mode estimates, afterwards these estimates are used as starting
values for the MCMC chains.
The returned object is of class `"bamlss"` for which generic extractor functions like
`summary()`, `plot()`, `predict()`, etc., are provided. For example, the estimated effects
for distribution paramaters `mu` and `sigma` can be visualized by
```{r, eval=FALSE}
plot(b, model = c("mu", "sigma"))
```
```{r, fig.width = 9, fig.height = 3.5, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE}
par(mar = c(4.1, 4.1, 1.1, 1.1), mfrow = c(1, 2))
plot(b, pages = 1, spar = FALSE, scheme = 2, grid = 100)
```
The model summary gives
```{r}
summary(b)
```
showing, e.g., the acceptance probabilities of the MCMC chains (`alpha`), the estimated degrees of freedom of
the optimizer and the successive sampler (`edf`), the final AIC and DIC as well as parametric
model coefficients (in this case only the intercepts). Using MCMC involves convergence checks of the
sampled parameters. The easiest diagnostics are traceplots
```{r, eval=FALSE}
plot(b, which = "samples")
```
```{r, fig.width = 9, fig.height = 5, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE}
bsp <- b
bsp$samples <- bsp$samples[, c("mu.p.(Intercept)", "sigma.p.(Intercept)")]
plot(bsp, which = "samples")
```
Note that this call would show all traceplots, for convenience we only show the plots for the intercepts.
In this case, the traceplots indicate convergence of the Markov chains (only the ACF for parameter `"mu"`
still shows some autocorrelation, which could be further reduced, e.g., by increasing the thinning
parameter). Further inspections are the maximum autocorrelation of all parameters, `which = "max-acf"`,
besides other convergence diagnostics, e.g., diagnostics that are part of the _coda_ package
[@bamlss:Plummer+Best+Cowles+Vines:2006].

Inspecting randomized quantile residuals [@bamlss:Dunn+Gordon:1996] is useful for judging how
well the model fits to the data
```{r, eval = FALSE}
plot(b, which = c("hist-resid", "qq-resid"))
```
```{r, fig.width = 8, fig.height = 4, fig.align = "center", echo = FALSE, dev = "png", echo=FALSE}
par(mfrow = c(1, 2))
plot(b, which = c("hist-resid", "qq-resid"), spar = FALSE)
```
Randomized quantile residuals are the default method in _bamlss_, which are computed using
the cdf function of the corresponding family object.

The posterior mean function for new data based on MCMC samples for parameter $\mu$ can be computed by
```{r, eval=FALSE}
nd <- data.frame("times" = seq(2.4, 57.6, length = 100))
nd$p <- predict(b, newdata = nd, model = "mu", FUN = mean)
plot2d(p ~ times, data = nd)
```
```{r, fig.width = 7, fig.height = 4, fig.align = "center", dev = "png", results='hide', message=FALSE, echo=FALSE, out.width="60%"}
par(mar = c(4.1, 4.1, 1.1, 1.1))
nd <- data.frame("times" = seq(2.4, 57.6, length = 100))
nd$p <- predict(b, newdata = nd, model = "mu", FUN = mean)
plot2d(p ~ times, data = nd)
```
where argument `FUN` can be any function, e.g., a function computing credible intervals from
the empirical quantiles of the MCMC samples
```{r}
foo <- function(x) {
  quantile(x, probs = c(0.025, 0.5, 0.975))
}
nd$p <- predict(b, newdata = nd, model = "mu", FUN = foo)
print(head(nd))
```

## References

