---
title: "Distributional Regression with _bamlss_"
author: "Nikolaus Umlauf, Nadja Klein, Achim Zeileis, Thorsten Simon"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{First Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{distributional regression, first steps}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
load("figures/toymodel.rda")
```

## Overview

The R package _bamlss_ provides a modular computational framework for distributional regression
models (and beyond). The implementation follows the conceptional framework presented in
[@bamlss:Umlauf+Klein+Zeileis:2017], which supports Bayesian and/or frequentist estimation engines
using complex possibly nonlinear model terms of any type. The highlights of the package are:

* A unified model description where a `formula` specifies how to set up the predictors
  from the `data` and `family` objects provide information about the response distribution.
* A generic method for setting up model terms and a `model.frame` along with the
  corresponding prior structures. A `transformer` can optionally set up modified terms, e.g.,
  using mixed model representation for smooth terms.
* Support for modular and exchangeable updating functions or complete model fitting engines
  in order to optionally implement either algorithms for maximization of the log-posterior for
  posterior mode estimation or for solving high-dimensional integrals, e.g., for posterior mean
  or median estimation.
  First, an (optional) `optimizer` function can be run, e.g., for computing posterior mode
  estimates. Second, a `sampler` is employed for full Bayesian inference with MCMC, which uses the
  posterior mode estimates from the `optimizer` as staring values. An additional step can be used
  for preparing the `results`.
* Standard post-modeling extractor functions to create sampling statistics, visualizations,
  predictions, etc.

## Installation

The stable release version of _colorspace_ is hosted on the Comprehensive R Archive Network
(CRAN) at <https://CRAN.R-project.org/package=bamlss> and can be installed via

```{r installation-cran, eval=FALSE}
install.packages("bamlss")
```

The development version of _bamlss_ is hosted on R-Forge at
<https://R-Forge.R-project.org/projects/bayesr/> in a Subversion (SVN) repository.
It can be installed via

```{r installation-rforge, eval=FALSE}
install.packages("bamlss", repos = "http://R-Forge.R-project.org")
```

## Running your first model

As a startup on how to use _bamlss_, we illustrate the basic steps on a small textbook
example using the well-known simulated motorcycle accident data [@bamlss:Silverman:1985].
The data contain measurements of the head acceleration
(in $g$, variable `accel`) in a simulated motorcycle accident, recorded in milliseconds after
impact (variable `times`). To estimate a Gaussian location-scale model with
$$
\texttt{accel} \sim \mathcal{N}(\mu = f(\texttt{times}), \log(\sigma) = f(\texttt{times}))
$$
we use the following model formula
```{r}
f <- list(accel ~ s(times, k = 20), sigma ~ s(times, k = 20))
```
where `s()` is the smooth term constructor from the _mgcv_ [@bamlss:Wood:2018]. Note,
that formulae are provided as `list`s of formulae, i.e., each list entry represents one
parameter of the response distribution. Also note that all
smooth terms, i.e., `te()`, `ti()`, etc., are supported by _bamlss_. This way, it is also
possible to incorporate user defined model terms. A full Bayesian model is the
estimated with
```{r, eval=FALSE}
library("bamlss")
data("mcycle", package = "MASS")

b <- bamlss(f, data = mcycle, family = "gaussian")
```
The returned object is of class `"bamlss"` for which generic extractor functions like
`summary()`, `plot()`, `predict()`, etc., are provided. For example, the estimated effects
for distribution paramaters `mu` and `sigma` can be visualized by
```{r, eval=FALSE}
plot(b, model = c("mu", "sigma"))
```
```{r, fig.width = 9, fig.height = 3.5, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE}
par(mar = c(4.1, 4.1, 1.1, 1.1), mfrow = c(1, 2))
plot(b, pages = 1, spar = FALSE, scheme = 2, grid = 100)
```
The model summary gives
```{r}
summary(b)
```

## References
