---
title: "BAMLSS Model Frame"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{BAMLSS model frame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{model frame}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
set.seed(123)
```

## Understanding the BAMLSS model frame

Similar to the well-known `model.frame()` function that is used, e.g., by the linear model 
fitting function `lm()`, or for generalized linear models `glm()`, the `bamlss.frame()`
function extracts a ``model frame'' for fitting distributional regression models. 
Internally, the function parses model formulae, one for each parameter of the distribution,
using the _Formula_ package infrastructures [@bamlss:Zeileis+Croissant:2010] in combination
with `model.matrix()` processing for linear effects and `smooth.construct()` processing of
the _mgcv_ package to setup design and penalty matrices for unspecified smooth function 
estimation [@bamlss:Wood:2018] (see also, e.g., the documentation of function `s()` and
`te()`).

The most important arguments are
```{r, eval=FALSE}
bamlss.frame(formula, data = NULL, family = "gaussian",
  weights = NULL, subset = NULL, offset = NULL,
  na.action = na.omit, contrasts = NULL, ...)
```
which basically represent the standard model frame specifications
[@bamlss:Chambers+Hastie:1992], while argument `formula` can be a classical model
formulae, e.g., as used by the `lm()` function, or an extended _bamlss_ formula, that is
internally parsed by function `bamlss.formula()`. Note that the _bamlss_ package uses
special `family` objects, that can be passed either as a character without the `"_bamlss"`
extension of the _bamlss_ family name (see the manual `?bamlss.family` for a list of
families), or the family function itself. In addition,
all families of the _gamlss_ [@bamlss:gamlss] and _gamlss.dist_ 
[@bamlss:gamlss.dist] package are supported.

## References

