---
title: "BAMLSS Model Frame"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{BAMLSS model frame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{model frame}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
set.seed(123)
prefix <- "http://bayesr.R-Forge.R-project.org/articles/" ## ""
```

## Understanding the BAMLSS model frame

Similar to the well-known `model.frame()` function that is used, e.g., by the linear model 
fitting function `lm()`, or for generalized linear models `glm()`, the `bamlss.frame()`
function extracts a "model frame" for fitting distributional regression models. 
Internally, the function parses model formulae, one for each parameter of the distribution,
using the _Formula_ package infrastructures [@bamlss:Zeileis+Croissant:2010] in combination
with `model.matrix()` processing for linear effects and `smooth.construct()` processing of
the _mgcv_ package to setup design and penalty matrices for unspecified smooth function 
estimation [@bamlss:Wood:2019], see also, e.g., the documentation of function `s()` and
`te()`.

The most important arguments are
```{r, eval=FALSE}
bamlss.frame(formula, data = NULL, family = "gaussian",
  weights = NULL, subset = NULL, offset = NULL,
  na.action = na.omit, contrasts = NULL, ...)
```
The argument `formula` can be a classical model formulae, e.g., as used by the `lm()` function,
or an extended _bamlss_ formula including smooth term specifications like `s()` or `te()`,
that is internally parsed by function `bamlss.formula()`.
Note that the _bamlss_ package uses special `family` objects, that can be passed either as
a character without the `"_bamlss"` extension of the _bamlss_ family name (see the manual
`?bamlss.family` for a list of available families and the corresponding vignette
[BAMLSS Families](`r prefix`families.html)), or the family function itself.
In addition, all families of the _gamlss_ [@bamlss:Stasinopoulos+Rigby:2019] and
_gamlss.dist_ [@bamlss:gamlss.dist] package are supported.

The returned object, a named list of class `"bamlss.frame"`, can be employed with all model
fitting engines. The most important elements used for estimation are:

* `x`: A named list, the elements correspond to the parameters that are specified within
  the `family` object. For each distribution parameter, the list contains all design and
  penalty matrices needed for modeling (see the upcoming example).
* `y`: The response data.
* `family`: The processed \pkg{bamlss} \code{family}.

To better understand the structure of the `"bamlss.frame"` object a print method is provided.
For illustration, we simulate data
```{r}
set.seed(111)
d <- GAMart()
```
and set up a `"bamlss.frame"` object for a Gaussian
distributional regression model including smooth terms. First, a model formula is needed
```{r}
f <- list(
  num ~ x1 + s(x2) + s(x3) + te(lon,lat),
  sigma ~ x1 + s(x2) + s(x3) + te(lon,lat)
)
```
Afterwards the model frame can be computed with
```{r}
bf <- bamlss.frame(f, data = d, family = "gaussian")
```
To keep the overview, there is also an implemented print method for `"bamlss.frame"` objects.
```{r}
print(bf)
```
For writing a new estimation engine, the user can directly work with the `model.matrix`
elements, for linear effects, and the `smooth.construct` list, for smooth effects respectively.
The `smooth.construct` is a named list which is compiled using the `smoothCon()` function
of the _mgcv_ package using the generic `smooth.construct()` method for setting up
smooth terms.
```{r}
print(names(bf$x$mu$smooth.construct))
```
In this example, the list contains three smooth term objects for parameter `mu` and
`sigma`.

See also the vignette [Estimation Engines](`r prefix`engines.html) presenting more details
on how to work with the `bamlss.frame()`.

## References

