---
title: "Bivariate Gaussian models for wind vectors"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{Model Terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{bivariate Gaussian distribution, distributional regression, 
    wind vectors, ensemble forecasts}
  %\VignettePackage{bamlss}
---

```{r preliminaries, include=FALSE}
library("zoo") 
library("bamlss")

if (!dir.exists("figures")) dir.create("figures")

if(!file.exists("figures/bivnorm_models.rda")) {

  # -------------------------------------------------------------------
  # Formulas
  # -------------------------------------------------------------------
  ## Baseline model BLM-0 (w/o correlation)
  f_blm.u.0 <- list(u10m_obs ~ s(yday, bs = 'cc') + s(yday, bs = 'cc', by = u10m_mean),
                    sigma    ~ s(yday, bs = 'cc') + s(yday, bs = 'cc', by = u10m_sd.log))
                    
  f_blm.v.0 <- list(v10m_obs ~ s(yday, bs = 'cc') + s(yday, bs = 'cc', by = v10m_mean),
                    sigma    ~ s(yday, bs = 'cc') + s(yday, bs = 'cc', by = v10m_sd.log ))
                    
  ## Rotation allowing model RAM-0 (w/o correlation)
  f_ram.u.0 <- list(u10m_obs ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), 
                                 bs = c('cc','cc'), by = u10m_mean) +
                                 te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_mean), 
                    sigma    ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'),
                                 by = u10m_sd.log) +
                                 te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_sd.log)) 
                                 
  f_ram.v.0 <- list(v10m_obs ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'),
                                 by = v10m_mean) +
                                 te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_mean), 
                    sigma    ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'),
                                 by = v10m_sd.log) +
                                 te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_sd.log)) 
                                 
  ## Rotation allowing model RAM-ADV (w/ correlation)
  f_ram.adv <- list(u10m_obs ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'),
                                 by = u10m_mean) +
                                 te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_mean), 
                    v10m_obs ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'),
                                 by = v10m_mean) +
                                 te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_mean), 
                    sigma1   ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'),
                                 by = u10m_sd.log) +
                                 te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_sd.log), 
                    sigma2   ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'),
                                 by = v10m_sd.log) +
                                 te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_sd.log), 
                    rho      ~   s(yday, bs = 'cc') + s(dd10m_mean, bs = 'cc') + s(dd10m_mean, bs = 'cc',
                                 by = ff10m_mean))
                    
  # -------------------------------------------------------------------
  # Fitting
  # -------------------------------------------------------------------
  ## Load data
  dat <- readRDS("figures/bivnorm_data.rds")
  dat <- dat[["Innsbruck"]]
  
  ## Baseline model BLM-0 (w/o correlation)
  m_blm.u.0 <- bamlss(f_blm.u.0, family = bamlss:::gaussian_bamlss, data = dat, optimizer = bfit,
                 burnin = 2000, thin = 10, n.iter = 12000, sampler = TRUE, nu = 1)
                 
  m_blm.v.0 <- bamlss(f_blm.v.0, family = bamlss:::gaussian_bamlss, data = dat, optimizer = bfit,
                 burnin = 2000, thin = 10, n.iter = 12000, sampler = TRUE, nu = 1)
                 
  ## Rotation allowing model RAM-0 (w/o correlation)
  m_ram.u.0 <- bamlss(f_ram.u.0, family = bamlss:::gaussian_bamlss, data = dat, optimizer = bfit,
                 burnin = 2000, thin = 10, n.iter = 12000, sampler = TRUE, nu = 1)

  m_ram.v.0 <- bamlss(f_ram.v.0, family = bamlss:::gaussian_bamlss, data = dat, optimizer = bfit,
                 burnin = 2000, thin = 10, n.iter = 12000, sampler = TRUE, nu = 1)

  ## Rotation allowing model RAM-ADV (w/ correlation)
  m_ram.adv <- bamlss(f_ram.adv, family = bamlss:::bivnorm_bamlss, data = dat, optimizer = bfit,
                 burnin = 2000, thin = 10, n.iter = 12000, sampler = TRUE, nu = 1)

  save(m_blm.u.0, m_blm.v.0, m_ram.u.0, m_ram.v.0, m_ram.adv, file = "figures/bivnorm_models.rda")
} else {

  load(file = "figures/bivnorm_models.rda")

}

```

## Intro

The article **Bivariate Gaussian models for wind vectors in a distributional
regression framework** [@bamlss:Lang+Mayr+Stauffer+Zeileis:2019] presents a new
probabilistic post-processing method for wind vectors employing the bivariate
Gaussian distribution [@bamlss:Klein+Kneib+Klasen+Lang:2014]. In contrast to
previous studies all parameters of the bivariate Gaussian distribution can be
modeled simultaneously, namely the location and scale parameters for both wind
components and also the correlation parameter between them employing flexible
regression splines. The performance of the new method is tested in
@bamlss:Lang+Mayr+Stauffer+Zeileis:2019 for stations located in plains,
mountain foreland, and within an alpine valley employing European Centre for
Medium-Range Weather Forecasts (ECMWF) ensemble forecasts as explanatory
variables for all distribution parameters.

This vignette provides example code how to set up the models employing the
package _bamlss_ as presented in the article.

## Data

The example code for the bivariate Gaussian models introduced in
@bamlss:Lang+Mayr+Stauffer+Zeileis:2019 is illustrated based on the freely
available EPS reforecasts of the NOAA's second-generation global ensemble
reforecast dataset [@bamlss:hamill+etal:2013] for the 12 hour forecast step
bilinearly interpolated to the site of interest. As observational data wind
measurement at the [Department of Atmospheric and Cryospheric
Sciences](http://acinn.uibk.ac.at/current-weather) of the University of
Innsbruck are used in this analysis ([CC BY-SA
4.0](http://creativecommons.org/licenses/by-sa/4.0/)). In addition, wind
measurements for Hamburg given by the Deutscher Wetterdienst
([DWD](https://www.dwd.de)) and the corresponding ensemble forecasts are also
provided but not analyzed in this vignette. At both sites, wind speed and
direction measurements are reported for the 10m height level. The data are
10-minute averages and contain the period from January 1, 2015 to December 31,
2018.

The data for Innsbruck and Hamburg is provided
[here](http://bayesr.r-forge.r-project.org/misc/bivnorm_data.rds) and can be
loaded into R with

```{r, eval = FALSE}
library("zoo")
dat <- readRDS("bivnorm_data.rds")
dat <- dat[["Innsbruck"]]
```

## Baseline model without correlation

In a first step, we combine two univariate heteroscedastic regression models
that post-process each wind component separately with correlation fixed at
zero. Hence, for the location and scale part, it uses its direct counterparts
of the EPS as covariates, namely EPS-forecasted zonal wind
information ($\text{vec}_1$) to model the zonal component of the bivariate
response, and EPS-forecasted meridional wind information ($\text{vec}_2$) to
model the meridional component:

\begin{align}
  \begin{split}
    \mu_{\star} &=
      \underbrace{{\alpha_\star}_{0} + {f_\star}_{0}(\text{doy})}_\text{intercept} +
      \underbrace{({\alpha_\star}_{1} + {f_\star}_{1}(\text{doy}))}_\text{slope coefficient}
      \cdot \text{vec}_{\star, mean},\\[1ex]
    \text{log}(\sigma_{\star}) &=
      \underbrace{{\beta_\star}_{0} + {g_\star}_{0}(\text{doy})}_\text{intercept} +
      \underbrace{({\beta_\star}_{1} + {g_\star}_{1}(\text{doy}))}_\text{slope coefficient}
      \cdot \text{vec}_{\star, log.sd},\label{eq:gam:locsc:v1}
  \end{split}
\end{align}

where, $\alpha_\bullet$ and $\beta_\bullet$ are regression coefficients, and
$f_\bullet(\text{doy})$ and $g_\bullet(\text{doy})$ employ cyclic regression
splines conditional on the day of the year (doy). The subscripts $mean$ and
$log.sd$ refer to mean and log standard deviation of the ensemble wind
components, respectively. The equation specifies a time-adaptive training
scheme, where the linear predictors consist of a global intercept and slope
coefficient plus a seasonally varying deviation. Thus, the intercept and slope
coefficients can smoothly evolve over the year in case that the bias or the
covariate's skill varies seasonally.

The formulas, model fitting, and prediction are performed according to:

```{r, eval=FALSE}
## Formulas
f_blm.u.0 <- list(u10m_obs ~ s(yday, bs = 'cc') + s(yday, bs = 'cc', by = u10m_mean),
                  sigma    ~ s(yday, bs = 'cc') + s(yday, bs = 'cc', by = u10m_sd.log))

f_blm.v.0 <- list(v10m_obs ~ s(yday, bs = 'cc') + s(yday, bs = 'cc', by = v10m_mean),
                  sigma    ~ s(yday, bs = 'cc') + s(yday, bs = 'cc', by = v10m_sd.log ))

## Model fitting
m_blm.u.0 <- bamlss(f_blm.u.0, family = bamlss:::gaussian_bamlss, data = dat, optimizer = bfit,
               burnin = 2000, thin = 10, n.iter = 12000, sampler = TRUE, nu = 1)
               
m_blm.v.0 <- bamlss(f_blm.v.0, family = bamlss:::gaussian_bamlss, data = dat, optimizer = bfit,
               burnin = 2000, thin = 10, n.iter = 12000, sampler = TRUE, nu = 1)

## Prediction
p_blm.0 <- list("mu1" = predict(m_blm.u.0, type = "parameter")$mu,
                "mu2" = predict(m_blm.v.0, type = "parameter")$mu,
                "sigma1" = predict(m_blm.u.0, type = "parameter")$sigma,
                "sigma2" = predict(m_blm.v.0, type = "parameter")$sigma,
                "rho" = rep(0, unique(sapply(predict(m_blm.v.0, type = "parameter"), length))))
```
The estimated cyclic seasonal effects are visualized for the first wind component ($\text{vec}_1$) with

```{r, eval=FALSE}
plot(m_blm.u.0, pages = 1)
```

```{r, fig.width = 6, fig.height = 6, fig.align = "center", dev = "png", echo = FALSE, dev = "png", results = 'hide', message=FALSE}
par(mar = c(4.5, 4, 0.5, 0.5))
plot(m_blm.u.0, pages = 1)
```

For the second wind component ($\text{vec}_2$) the effects are illustrated accordingly: 

```{r, fig.width = 6, fig.height = 6, fig.align = "center", dev = "png", echo = FALSE, dev = "png", results = 'hide', message=FALSE}
par(mar = c(4.5, 4, 0.5, 0.5))
plot(m_blm.v.0, pages = 1)
```

The estimated effects are on the scale of the additive predictor; i.e., on the
linear scale for the location parameters $\mu_\star$ and on the log-scale for
the scale parameters $\sigma_\star$. Each of the four distribution parameters
is described by a (potentially) seasonally varying effect for the intercept and
the slope coefficient.

## Rotation-allowing model without correlation

In a second step, we extend the setup by employing the zonal and meridional wind
information of the ensemble for the linear predictors of all location and scale
parameters. That means we use the ensemble information of both the zonal and
meridional wind components for the two components of the response.
We propose to employ seasonally varying effects depending on the ensemble
wind direction, which allows the model to rotate the forecasted wind direction
if necessary. To do so, we obtain a two-dimensional smooth function represented
by a tensor product spline with a respective cyclic constraint for the day of
the year (doy) and for the mean ensemble wind direction ($\text{dir}_{mean}$):

\begin{equation}
  \begin{split} 
    \mu_{\star} = {\alpha_\star}_{0} + {f_\star}_{0}(\text{doy}) &+  
      ({\alpha_\star}_{1} + {f_\star}_{1}(\text{doy}) \cdot {f_\star}_{2}(\text{dir}_{mean})) \cdot \text{vec}_{1, mean}\\
      &+ ({\alpha_\star}_{2} + {f_\star}_{3}(\text{doy}) \cdot {f_\star}_{4}(\text{dir}_{mean})) \cdot \text{vec}_{2, mean},\\[1ex]
    \text{log}(\sigma_{\star}) = {\beta_\star}_{0} + {g_\star}_{0}(\text{doy}) &+  
      ({\beta_\star}_{1} + {g_\star}_{1}(\text{doy}) \cdot {g_\star}_{2}(\text{dir}_{mean})) \cdot \text{vec}_{1, log.sd}\\
      &+ ({\beta_\star}_{2} + {g_\star}_{3}(\text{doy}) \cdot {g_\star}_{4}(\text{dir}_{mean})) \cdot \text{vec}_{2, log.sd},
    \label{eq:gam:locsc:v2}
  \end{split}
\end{equation}

where, as before, $\alpha_\bullet$ and $\beta_\bullet$ are regression
coefficients, and $f_\bullet$ and $g_\bullet$ employ cyclic regression splines.

The model formulas are specified by

```{r, eval=FALSE}
## Formulas
f_ram.u.0 <- list(u10m_obs ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_mean) +
                               te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_mean),
                  sigma    ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_sd.log) +
                               te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_sd.log))

f_ram.v.0 <- list(v10m_obs ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_mean) +
                               te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_mean),
                  sigma    ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_sd.log) +
                               te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_sd.log))
```

## Rotation-allowing models with correlation

We further extend the setup by explicitly modeling the correlation parameter in
a bivariate Gaussian setup. For the estimation of the correlation structure, we
assume that the correlation mainly depends on the mean ensemble wind direction
($\text{dir}_{mean}$) and speed ($\text{spd}_{mean}$) by modeling a linear
interaction between these two covariates:

\begin{align}
\text{rhogit}(\rho) &= \gamma_0 + h_0(\text{doy}) +
  h_1(\text{dir}_{mean}) +
  (\gamma_1 + h_2(\text{dir}_{mean})) \cdot \text{spd}_{mean}\label{eq:gam:rho},
\end{align}

with $\text{rhogit}(\rho) = \rho / \sqrt{(1 - \rho^2)}$; $\gamma_0$ is the
global intercept and $h_0(\text{doy})$ the seasonally varying intercept. The
effect $h_1(\text{dir}_{mean})$ estimates the dependence of the correlation
given the wind direction and $(\gamma_1 + h_2(\text{dir}_{mean})) \cdot
\text{spd}_{mean}$ employs a varying effect of wind speed conditional on the
wind direction. 

The formulas, model fitting, and prediction are performed according to:

```{r, eval=FALSE}
## Formulas
f_ram.adv <- list(u10m_obs ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_mean) +
                               te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_mean), 
                  v10m_obs ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_mean) +
                               te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_mean), 
                  sigma1   ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_sd.log) +
                               te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_sd.log), 
                  sigma2   ~ s(yday, bs = 'cc') + te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = v10m_sd.log) +
                               te(yday, dd10m_mean, k = c(5, 5), bs = c('cc','cc'), by = u10m_sd.log),
                  rho      ~   s(yday, bs = 'cc') + s(dd10m_mean, bs = 'cc') + s(dd10m_mean, bs = 'cc', by = ff10m_mean))

## Model fitting
m_ram.adv <- bamlss(f_ram.adv, family = bamlss:::bivnorm_bamlss, data = dat, optimizer = bfit,
               burnin = 2000, thin = 10, n.iter = 12000, sampler = TRUE, nu = 1)

## Prediction
p_ram.adv <- predict(m_ram.adv, type = "parameter")
```

The distribution of the correlation parameters for the underlying dependence
structure of the raw ensemble and for the fitted bivariate model is visualized with

```{r, eval=FALSE}
hist(dat$uv10m_cor, xlim = c(-1,1), col = gray(0.8), freq = FALSE, breaks = 17, 
  xlab = "Correlation parameter", main = "Raw ensemble", cex.main = 0.8)

hist(p_ram.adv$rho, xlim = c(-1,1), col = gray(0.8), freq = FALSE, breaks = 17, 
  xlab = "Correlation parameter", main = "Fitted bivariate model", cex.main = 0.8)
```

```{r, fig.width = 8, fig.height = 4, fig.align = "center", echo = FALSE, dev = "png"}
par(mfrow = c(1, 2))

dat <- readRDS("figures/bivnorm_data.rds")
dat <- dat[["Innsbruck"]]
hist(dat$uv10m_cor, xlim = c(-1,1), col = gray(0.8), freq = FALSE, breaks = 17, 
  xlab = "Correlation parameter", main = "Raw ensemble", cex.main = 0.8)

suppressWarnings(
  p_ram.adv <- predict(m_ram.adv, type = "parameter")
)
hist(p_ram.adv$rho, xlim = c(-1,1), col = gray(0.8), freq = FALSE, breaks = 17, 
  xlab = "Correlation parameter", main = "Fitted bivariate model", cex.main = 0.8)
```

## References
