---
title: "Complex Space-Time Interactions in a Cox Model"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{Cox}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{distributional regression, survival, cox model}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
set.seed(123)
load("figures/firemodel.rda")
firemodel$family <- cox_bamlss()
```

## Intro

In the article **BAMLSS: Bayesian Additive Models for Location, Scale, and Shape (and Beyond)**
[@bamlss:Umlauf+Klein+Zeileis:2017] an example based on the article of @bamlss:Taylor:2015 is
used to illustrate the BAMLSS framework with complex space-time interactions in a Cox model.

The example uses data from the _London Fire Brigade_
(LFB, [http://www.london-fire.gov.uk/](http://www.london-fire.gov.uk/)), which is one of largest
fire brigades  in the world. Each year, the LFB is called thousands of times, in most cases due to
dwelling fires. To prevent further damage or fatal casualties, a short arrival time is important,
i.e., the time it takes until a fire engine arrives at the scene after an emergency call has been
received. The aim of this analysis is to explore the drivers of arrival times.

## Model

The response times are analyzed within a survival context where the
hazard of an event (fire engine arriving) at time $t$ follows a relative risk model of the form
$$
\lambda(t) = \exp\left(\eta(t)\right) =  \exp\left( \eta_{\lambda}(t) + \eta_{\gamma} \right),
$$
i.e., a model for the instantaneous arrival rate conditional on the engine not having arrived before
time $t$. Here, the hazard function is assumed to depend on a time-varying predictor
$\eta_{\lambda}(t)$ and a time-constant predictor $\eta_{\gamma}$.

We set up a model with the time-constant predictor
$$
\eta_{\gamma} = \beta_0 + f_1(\texttt{fsintens}) + f_2(\texttt{daytime}) +
  f_3(\texttt{lon}, \texttt{lat}) + f_4(\texttt{daytime}, \texttt{lon}, \texttt{lat}),
$$
where $\beta_0$ is an intercept and function $f_1( \cdot )$ is the effect of fire station intensity
(`fsintens`, computed with a kernel density estimate of all fire stations in London). The other
variables represent: The time of the day (`daytime`), exact spatial coordinates of the fire
(`lon` and `lat`). Here function $f_4( \cdot )$ is a three-dimensional interaction effect.

The time-varying additive predictor has the following form
$$
\eta_{\lambda}(\texttt{arrivaltime}) = f_0(\texttt{arrivaltime}) + f_1(\texttt{arrivaltime}, \texttt{lon}, \texttt{lat}),
$$
where $f_0( \cdot )$ is the baseline hazard for variable `arrivaltime`, the waiting time until the
first fire engine arrives after the received emergency call.

The probability that the engine will arrive on the scene after time $t$ is described by the survival
function
\begin{equation} \label{eqn:surv}
S(t) = \mathrm{Prob}(T > t) = \exp \, \left( -\int_0^t \lambda(u)du \right),
\end{equation}
which is of prime interest in this analysis.

## Data

The data is freely available from the London DataStore
([http://data.london.gov.uk/](http://data.london.gov.uk/)) under the UK
Open Government Licence (OGL v2). It can be downloaded from
[http://data.london.gov.uk/dataset/london-fire-brigade-incident-records](http://data.london.gov.uk/dataset/london-fire-brigade-incident-records)
which also contains previous years. The preprocessed data is part of the _bamlss_ package and
can be loaded with
```{r}
data(LondonFire, package = "bamlss")
head(LondonFire)
```
and is stored as a `"SpatialPointsDataFrame"`. Therefore, the spatial distribution of fires in
London along with boundary polygons and station locations can be plotted instantly with:
```{r, fig.width = 6, fig.height = 5, fig.align = "center", dev = "png", out.width="60%"}
library("sp")
par(mar = rep(0, 4))
plot(LondonBoundaries)
plot(LondonFire, col = "red", add = TRUE)
plot(LondonBoroughs, add = TRUE)
plot(LondonFStations, col = "blue", add = TRUE, pch = 16)
```

## Estimation

The Cox model is implemented in the `cox_bamlss()` family and uses a special optimizer function
`cox_mode()` and sampling engine `cox_mcmc()`. Note that the optimizer and sampler function do
not need to be called explicitly within the `bamlss()` wrapper call, because the `cox_bamlss()`
family specifies this already in its return value, such that internally function `bamlss()`
nows to only use the estimation engines that are supplied by the family.
```{r}
fam <- cox_bamlss()
names(fam)
```
There is also an additional transformer function, which is needed for computing the (numerical)
integrals that are part of the log-likelihood. Predictions are also based on integrals, therefore
the `predict()` function is also part of this family object and will be used by `predict.bamlss()`
instead of the default methods.

The model formula has two parts, the time-dependent for $\eta_{\lambda}$ and the time-constant
part for $\eta_{\gamma}$ and can be set up with
```{r}
f <- list(
  Surv(arrivaltime) ~ ti(arrivaltime,k=20) + ti(arrivaltime,lon,lat,d=c(1,2),k=c(5,30)),
  gamma ~ s(fsintens) + ti(daytime,bs="cc",k=30) + ti(lon,lat,k=80,d=2) +
    ti(daytime,lon,lat,bs=c("cc","cr"),d=c(1,2),k=c(10,30))
)
```
Note that the `Surv()` function from the _survival_ package [@bamlss:Therneau:2016] is used to
set up the formula for $\eta_{\lambda}$. Also note that the arrival times are not censored
in this application. The model is estimated with
```{r, eval=FALSE}
## Set the seed for reproducibility.
set.seed(222)

## Start estimation
firemodel <- bamlss(f, data = LondonFire, family = "cox",
  subdivisions = 25, maxit = 1000,
  n.iter = 6000, burnin = 3000, thin = 20, cores = 8)
```
Note, due to the complexity of the model and model terms estimation takes quite long. On a Linux
system with 8 Intel i7-2600 3.40GHz processors estimation takes approximately 1.2 days.

Good practice after fitting the model is to do some convergence checks of the MCMC chains, e.g.,
by looking at traceplots
```{r, eval=FALSE}
plot(firemodel, which = "samples")
```
```{r, fig.width = 9.5, fig.height = 4, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE, out.width="80%"}
bsp <- firemodel
bsp$samples <- bsp$samples[[1]][, "gamma.p.model.matrix.(Intercept)", drop = FALSE]
colnames(bsp$samples) <- "gamma.p.(Intercept)"
plot(bsp, which = "samples")
```
Note, for convenience we only show the traceplot of the intercept term of predictor $\eta_{\gamma}$.
The traceplot indicates convergence of the MCMC chains and have close to i.i.d. behavior.
The model summary gives
```{r}
summary(firemodel)
```
and indicates that the three-dimensional effects in `lambda` and `gamma` have an effect on the
response times. The estimated effects can be plotted with
```{r, eval=FALSE}
par(mar = c(4.5, 4, 0.5, 0.5), mfrow = c(2, 2))
plot(firemodel, scale = 0)
```
```{r, fig.width = 7.2, fig.height = 5.8, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE, out.width="80%"}
par(mar = c(4.1, 4.1, 1.1, 1.1), mfrow = c(2, 2))
plot(firemodel, scale = 0)
```
The upper left plot shows the estimated baseline hazard effect. The upper right plot the time-constant
effect of fire station intensity. The lower left plot the estimated effect of the time of the day
and the lower right plot the estimated time-constant spatial effect.

## Prediction

Predictions in for an estimated Cox model are based on the custom predict function `cox_predict()`.
Here, predicted probabilities are based on numerical integration, therefore, the user can specify
the subdivisions that are used in the integration routine. For example, let's pick a sample
location within the boundaries of London and predict the corresponding probability that the
fire engine arrives at 15pm within 0 to 20 minutes.
```{r}
## Extract the 150th sample.
i <- 150

## Create a new data frame for prediction.
nd <- data.frame(
  "arrivaltime" = seq(0, 20, length = 100),
  "daytime" = 15
)
nd$fsintens <- LondonFire$fsintens[i]
nd$lon <- LondonFire$lon[i]
nd$lat <- LondonFire$lat[i]

## Predict probabilities.
nd$p <- 1 - t(predict(firemodel, newdata = nd,
  type = "probabilities", subdivisions = 100, FUN = c95))
```
The estimated probabilities that the fire engine arrives after $t$ minutes can then be plotted with
```{r, fig.width = 5, fig.height = 4, fig.align = "center", dev = "png", out.width="50%"}
par(mar = c(4.1, 4.1, 0.1, 0.1))
plot2d(p ~ arrivaltime, data = nd, ylab = "1 - Prob(T > t)")
```

## References

