---
title: "Visualization with distreg.vis"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{Visualizing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{distributional regression, visualization}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
library("ggplot2")
options(width = 90, prompt = "R> ", continue = "+  ")
set.seed(123)
```

## R package distreg.vis

The [_distreg.vis_](https://github.com/Stan125/distreg.vis) [@bamlss:distreg.vis] package for R
allows us to interactively explore the predicted response distribution using different
covariate scenarios.

## Example: ISLR::Wage dataset
As a short introduction to distributional regression, we are going to take a look at a dataset on
the wage of 3000 male workers in the Mid-Atlantic Region of the US. The dataset is contained in the
[_ISLR_](https://cran.r-project.org/package=ISLR) [@bamlss:ISLR] package for R. We may output the
first 6 rows of the dataset with the following R command:

```{r}
head(ISLR::Wage)
```

The variable names and factor labels are mostly self-explanatory, but we can look up the details by
running `?ISLR::Wage`. Our goal is to link wage to a number of relevant explanatory variables. One
main driver of wage should be education, so let us start by descriptively investigating the
relationship between these two variables. This R code plots the distribution of wage across the
different education levels:

```{r, fig.width = 8.5, fig.height = 4, fig.align = "center", dev = "png"}
library("ggplot2")

ggplot(ISLR::Wage, aes(x = wage, y = ..density.., fill = education)) +
  geom_density(alpha = 0.8) +
  theme_classic() +
  ggtitle("Wages by education level")
```

The graph shows that education does have a big influence on wage. A higher education level
shifts the wage distribution to the right, but it also increases the variance and changes the
shape of the distribution. One striking feature of the empirical wage distribution is its bump
between 250 and 300. There is no explanation for this bump in the dataset or its help page, so we
have to assume it is a anomaly of the sample.

## Setting up an appropriate model

When modeling wage, we need to take into account its special characteristics as a random variable:
It is positive (or non-negative, but our dataset does not contain any zeros) and its distribution
is right-skewed. These characteristics make distributional regression an appropriate tool for the
task at hand.

Based on the characteristics of wage, we choose the
[log-normal distribution](https://en.wikipedia.org/wiki/Log-normal_distribution) for our model.
The log-normal distribution has positive support, it is right-skewed, and it has a location
parameter $\mu$ and a scale parameter $\sigma$. The idea of distributional regression is to link all
parameters of the response distribution to explanatory variables. In our case, these will be
education, age, race, and year.

The following R code specifies and estimates our model using the `LOGNO` family of the 
_gamlss_ package [@bamlss:Stasinopoulos+Rigby:2018]:

```{r, echo=FALSE, message=FALSE, results='hide'}
library("gamlss")

f <- list(
  wage ~ s(age) + race + year + education,
  sigma ~ s(age) + race + year + education
)

if(!file.exists("figures/distregvis.rda")) {
  set.seed(123)
  b <- bamlss(f, data = ISLR::Wage, family = LOGNO)
  save(b, file = "figures/distregvis.rda")
} else {
  load("figures/distregvis.rda")
}
```
```{r, eval=FALSE}
library("gamlss")

f <- list(
  wage ~ s(age) + race + year + education,
  sigma ~ s(age) + race + year + education
)

b <- bamlss(f, data = ISLR::Wage, family = LOGNO)
```

We may take a look at the estimation results with this R command:

```{r}
summary(b)
```

The output is similar to the summary of a linear model. For each parameter of the response
distribution, we get a regression table with slope coefficients, etc. Unfortunately, the
coefficients are hard to interpret because they are related to different aspects of the
response distribution. Moreover, the scale parameter $\sigma$ is transformed with a log-link.

## Explore the model with _distreg.vis_

One way to understand the meaning of the coefficients is to define different "covariate scenarios"
and take a look at the corresponding predicted response distributions. The
[_distreg.vis_](https://github.com/Stan125/distreg.vis) [@bamlss:distreg.vis] package
for R allows us to do so interactively:

<a target="_blank" class="btn btn-primary" href="http://141.5.104.49:3838/distreg_app/" role="button">Explore the model with distreg.vis</a>

## References

