---
title: "Estimation Engines"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{engines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{optimization, MCMC, estimation}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
set.seed(123)
```

## Intro

For the estimation of model parameters the _bamlss_ package provides a flexible
infrastructure that allows the user to exchange both, optimization functions for
posterior mode estimation (penalized likelihood) and sampling functions for full
Bayesian Inference. This goes beyond the common modeling infrastructures that usually
allow to create new family objects, only, and can be too restrictive for a couple of applications.
Within the unified modeling approach in _bamlss_ it is relatively
straightforward to develop new estimation functions (algorithms), e.g., for implementing
new models or for efficiency reasons. Such estimation functions could also interface to
other software (see, e.g., the `JAGS()` sampling function). In the following,
the basic requirements for optimizer and sampling functions are described in detail.

## The basic setup

Estimation engines in _bamlss_ are usually based on the model frame setup function
`bamlss.frame()`, i.e., the functions all have a `x` argument, which contains all the
necessary model and penalty matrices, and a `y` argument, which is the response
(univariate or multivariate). In addition, an estimation engine usually has a `family`
argument, which specifies the model to be estimated. However, this is not a mandatory
argument, i.e., one could write an estimation function that is designed for one specific
problem, only.

The modeling setup is best explained by looking at the main estimation engines provided by _bamlss_.
The default optimizer using the `bamlss()` wrapper function is `bfit()`, which is a
backfitting routine. The most important arguments are
```{r, eval=FALSE}
bfit(x, y, family, start = NULL, weights = NULL, offset = NULL, ...)
```
The default sampling engine in _bamlss_ is `GMCMC()`, again the most important
arguments are
```{r, eval=FALSE}
GMCMC(x, y, family, start = NULL, weights = NULL, offset = NULL, ...)
```
So basically, the arguments of the optimizer and the sampling function are the same, the
main difference is the return value. In _bamlss_ optimizer functions usually return
a vector of estimated regression coefficients (parameters), while sampling functions 
return a matrix of parameter samples of class `"mcmc"` or `"mcmc.list"` (for details see the 
documentation of the _coda_ package).

Internally, what the optimizer or sampling function is actually processing is not important
for the `bamlss()` wrapper function as long as a vector or matrix of parameters is
returned. For optimizer functions the return value needs to be named list with an element
`"parameters"`, the vector (also a matrix, e.g., for `lasso()` and `boost()` optimizers)
of estimated parameters. The most important requirement to make use of all extractor
functions like `summary.bamlss()`, `predict.bamlss()`, `plot.bamlss()`,
`residuals.bamlss()`, etc., is to follow the naming convention of the returned estimates.
The parameter names are based on the names of the distribution parameters as specified in
the family object. For example, the family object `gaussian_bamlss()` has parameter names `"mu"`
and `"sigma"`
```{r}
gaussian_bamlss()$names
```
Then, each distributional parameter can be modeled by parametric (linear) and
nonlinear smooth effect terms. The parametric part is indicated with `"p"` and the smooth part with
`"s"`. The names of the parametric coefficients are the names of the corresponding model matrices as
returned from `bamlss.frame()`. E.g., if two linear effects, "`x1`" and `"x2"`, enter the model for 
distributional parameter `"mu"`, then the final names are `"mu.p.x1"` and `"mu.p.x2"`.
Similarly for the smooth parts, if we model a variable `"x3"` using a regression spline as
provided by the `s()` function of the _mgcv_ package, the
name is based on the names that are used by `bamlss.frame()` for the `smooth.construct()`
object. In this case the parameter names start with `"mu.s.s(x3)"`. If this smooth
term has 10 regression coefficients, then the final name must be
```{r}
paste0("mu.s.s(x3)", ".b", 1:10)
```
i.e., all smooth term parameters are named with "b" and a numerated.

## Model fitting engines for linear regression

In the following, to explain the setup and the naming convention of estimation engines in
more detail, we implement

* a new family object for simple linear models $y = x^{\top}\boldsymbol{\beta} + \varepsilon$ with
  $\varepsilon \sim N(0, \sigma^2)$,
* and set up an optimizer function based on the `lm()` function,
* and additionally a MCMC sampling function.

For illustration, the family object is kept very simple, we only model the mean function 
in terms of covariates.
```{r}
lm_bamlss <- function(...) {
  f <- list(
    "family" = "LM",
    "names" = "mu",
    "links" = "identity",
    "d" = function(y, par, log = FALSE) {
      sigma <- sqrt(sum((y - par$mu)^2) / (length(y) - .lm_bamlss.p))
      dnorm(y, mean = par$mu, sd = sigma, log = log)
    },
    "p" = function(y, par, ...) {
      sigma <- sqrt(sum((y - par$mu)^2) / (length(y) - .lm_bamlss.p))
      pnorm(y, mean = par$mu, sd = sigma, ...)
    }
  )
  class(f) <- "family.bamlss"
  return(f)
}
```
Now, for setting up the estimation functions we first simulate some data using the
`GAMart()` function, afterwards the necessary `"bamlss.frame"` can be created with
```{r}
## Simulate some data.
d <- GAMart()

## Setup a "bamlss.frame" object, that is used for
## developing the estimation functions for the linear model.
bf <- bamlss.frame(num ~ x1 + x2, data = d, family = "lm")

## Print the structure of the "bamlss.frame".
print(bf)
```
As noted above, the object is a named list with elements `"x"` and `"y"`, which will be
passed to the estimation functions. For the moment, since we only implement a linear 
model, we need to work with the linear model matrix that is part of the `bf` object.
```{r}
print(head(bf$x$mu$model.matrix))
```
and the response `"y"`
```{r}
print(head(bf$y))
```
to setup the optimizer function with:
```{r}
## Linear model optimizer function.
lm.opt <- function(x, y, ...)
{
  ## Only univariate response.
  y <- y[[1L]]

  ## For illustration this is easier to read.
  X <- x$mu$model.matrix

  ## Estimate model parameters.
  par <- drop(chol2inv(chol(crossprod(X))) %*% crossprod(X, y))

  ## Set parameter names.
  names(par) <- paste0("mu.p.", colnames(X))

  ## Return estimated parameters and fitted values.
  rval <- list(
    "parameters" = par,
    "fitted.values" = drop(X %*% par),
    "edf" = length(par),
    "sigma" = drop(sqrt(crossprod(y - X %*% par) / (length(y) - ncol(X))))
  )

  ## Set edf within .GlobalEnv for the
  ## loglik() function in the lm_bamlss() family.
  .lm_bamlss.p <<- length(par)

  return(rval)
}
```
This optimizer function can already be used with the `bamlss()` wrapper function and
all extractor functions are readily available.
```{r}
## Model formula with polynomial model terms.
f <- num ~ x1 + poly(x2, 5) + poly(x3, 5)

## Estimate model with new optimizer function.
b <- bamlss(f, data = d, family = "lm", optimizer = lm.opt, sampler = FALSE)

## Summary output.
summary(b)

## Predict for term x2.
nd <- data.frame("x2" = seq(0, 1, length = 100))
nd$p <- predict(b, newdata = nd, term = "x2")
```
Plot the estimated effect of `x2`.
```{r, eval = FALSE}
plot2d(p ~ x2, data = nd)
```
```{r, fig.width = 5, fig.height = 4, fig.align = "center", dev = "png", echo = FALSE}
par(mar = c(4.1, 4.1, 0.1, 0.1))
plot2d(p ~ x2, data = nd)
```

The next step is to setup a full Bayesian MCMC sampling function. Fortunately, if we
assume multivariate normal priors for the regression coefficients and an inverse Gamma prior
for the variance, a Gibbs sampler with multivariate normal and inverse Gamma full conditionals
can be created. The MCMC algorithm consecutively samples for $t = 1, \ldots, T$ from the full
conditionals
$$
\boldsymbol{\beta}^{(t)} | \cdot \sim N\left(\boldsymbol{\mu}_{\boldsymbol{\beta}}^{(t - 1)}, 
  \boldsymbol{\Sigma}_{\boldsymbol{\beta}}^{(t - 1)}\right)
$$
and
$$ 
{\sigma^2}^{(t)} | \cdot \sim IG\left({a^{\prime}}^{(t - 1)}, {b^{\prime}}^{(t - 1)}\right),
$$
where $IG( \cdot )$ is the inverse Gamma distribution for sampling the variance parameter.
The covariance matrix for $\boldsymbol{\beta}$ is given by
$$
\boldsymbol{\Sigma}_{\boldsymbol{\beta}} = \left(\frac{1}{\sigma^2}\mathbf{X}^\top\mathbf{X} +
  \frac{1}{\sigma^2}\mathbf{M}^{-1}\right)^{-1}
$$
and the mean
$$
\boldsymbol{\mu}_{\boldsymbol{\beta}} = \boldsymbol{\Sigma}_{\boldsymbol{\beta}}
  \left(\frac{1}{\sigma^2}\mathbf{X}^\top\mathbf{y} +
  \frac{1}{\sigma^2}\mathbf{M}^{-1}\mathbf{m}\right),
$$
where $\mathbf{m}$ is the prior mean and $\mathbf{M}$ the prior covariance matrix.
Similarly, for $\sigma^2$ paramaters $a^{\prime}$ and $b^{\prime}$ are computed by
$$
a^{\prime} = a + \frac{n}{2} + \frac{p}{2}
$$
and
$$
b^{\prime} = b + \frac{1}{2}(\mathbf{y} -
  \mathbf{X}\boldsymbol{\beta})^\top(\mathbf{y} - \mathbf{X}\boldsymbol{\beta}) +
  \frac{1}{2} (\boldsymbol{\beta} - \mathbf{m})^\top \mathbf{M}^{-1}(\boldsymbol{\beta}
  - \mathbf{m}),
$$
where $a$ and $b$ are usually set small, e.g., with $a = 1$ and $b = 0.0001$, such that the prior is
flat and uninformative.

We can implement the MCMC algorithm in the following sampling function
```{r}
lm.mcmc <- function(x, y, start = NULL,
  n.iter = 12000, burnin = 2000, thin = 10,
  m = 0, M = 1e+05,
  a = 1, b = 1e-05,
  verbose = TRUE, ...)
{
  ## How many samples are saved?
  itrthin <- seq.int(burnin, n.iter, by = thin)
  nsaves <- length(itrthin)

  ## Only univariate response.
  y <- y[[1L]]

  ## For illustration this is easier to read.
  X <- x$mu$model.matrix

  ## Again, set edf within .GlobalEnv for the
  ## loglik() function in the lm_bamlss() family.
  .lm_bamlss.p <<- ncol(X)

  ## Number of observations and parameters.
  n <- length(y)
  p <- ncol(X)

  ## Matrix saving the samples.
  samples <- matrix(0, nsaves, p + 1L)

  ## Stick to the naming convention.
  pn <- paste0("mu.p.", colnames(X))
  colnames(samples) <- c(
    pn,      ## Regression coefficients and
    "sigma"  ## variance samples.
  )

  ## Setup coefficient vector,
  ## again, use correct names.
  beta <- rep(0, p)
  names(beta) <- pn
  sigma <- sd(y)

  ## Check for starting values obtained,
  ## e.g., from lm.opt() from above.
  if(!is.null(start)) {
    sn <- names(start)
    for(j in names(beta)) {
      if(j %in% sn)
        beta[j] <- start[j]
    }
  }

  ## Process prior information.
  m <- rep(m, length.out = p)
  if(length(M) < 2)
    M <- rep(M, length.out = p)
  if(!is.matrix(M))
    M <- diag(M)
  Mi <- solve(M)

  ## Precompute cross products.
  XX <- crossprod(X)
  Xy <- crossprod(X, y)

  ## Inverse gamma parameter.
  a <- a + n / 2 + p / 2

  ## Start sampling.
  ii <- 1
  for(i in 1:n.iter) {
    ## Sampling sigma
    b2 <- b + 1 / 2 * t(y - X %*% beta) %*% (y - X %*% beta) +
      1 / 2 * t(beta - m) %*% Mi %*% (beta - m)
    sigma2 <- sqrt(1 / rgamma(1, a, b2))

    ## Sampling beta.
    sigma2i <- 1 / sigma2
    Sigma <- chol2inv(chol(sigma2i * XX + sigma2i * Mi))
    mu <- Sigma %*% (sigma2i * Xy + sigma2i * Mi %*% m)
    beta <- MASS::mvrnorm(1, mu, Sigma)
      
    if(i %in% itrthin) {
      samples[ii, pn] <- beta
      samples[ii, "sigma"] <- sqrt(sigma2)
      ii <- ii + 1
    }
    if(verbose) {
      if(i %% 1000 == 0)
        cat("iteration:", i, "\n")
    }
  }

  ## Convert to "mcmc" object.
  samples <- as.mcmc(samples)

  return(samples)
}
```
The new estimation function can be directly used with the `bamlss()` wrapper.
```{r}
b <- bamlss(f, data = d, family = "lm", optimizer = lm.opt, sampler = lm.mcmc)

summary(b)

## Predict for all terms including 95% credible intervals
nd$x1 <- nd$x3 <- seq(0, 1, length = 100)
for(j in c("x1", "x2", "x3"))
  nd[[paste0("p.", j)]] <- predict(b, newdata = nd, term = j, FUN = c95, intercept = FALSE)
```
The estimated effects can be plotted with.
```{r, eval = FALSE}
par(mfrow = c(1, 3))
plot2d(p.x1 ~ x1, data = nd, fill.select = c(0, 1, 0, 1), lty = c(2, 1, 2))
plot2d(p.x2 ~ x2, data = nd, fill.select = c(0, 1, 0, 1), lty = c(2, 1, 2))
plot2d(p.x3 ~ x3, data = nd, fill.select = c(0, 1, 0, 1), lty = c(2, 1, 2))
```
```{r, fig.width = 10, fig.height = 4, fig.align = "center", dev = "png", echo = FALSE}
par(mar = c(4.1, 4.1, 0.5, 1.1), mfrow = c(1, 3))
plot2d(p.x1 ~ x1, data = nd, fill.select = c(0, 1, 0, 1), lty = c(2, 1, 2))
plot2d(p.x2 ~ x2, data = nd, fill.select = c(0, 1, 0, 1), lty = c(2, 1, 2))
plot2d(p.x3 ~ x3, data = nd, fill.select = c(0, 1, 0, 1), lty = c(2, 1, 2))
```

## References

