---
title: "BAMLSS Families"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{Families}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{distributional regression, families}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
set.seed(123)
```

## Overview

Family objects are important building blocks in the design of BAMLSS models.
They specify the distribution by collecting functions of the density,
respective log-likelihood, first-order derivatives of the log-likelihood w.r.t.
predictors, and (optionally) second-order derivatives of the log-likelihood
w.r.t. predictors.

## How to build a BAMLSS family

## Available families

There are some families implemented:
```{r}
x <- scan("../R/families.R", what = character(0))
x[grep("^\\w*_bamlss", x)]
```

### Continuous response

* `gaussian_bamlss`

### Positive continuous response

* `gpareto_bamlss`
* `weibull_bamlss`
* `gamma_bamlss`
* `lognormal_bamlss`

### Censored continuous responses

* `cnorm_bamlss`

### Interval responses

* `beta_bamlss`

### Discrete responses

* `binomial_bamlss`
* `mlt_bamlss`

### Count data responses

* `poisson_bamlss`
* `negbin_bamlss`

### Multivariate responses

* `bivnorm_bamlss`
* `mvnorm_bamlss`

### Others



