---
title: "Bayesian Joint Models for Longitudinal and Survival Data"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{LASSO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{distributional regression, joint models}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
set.seed(123)
```

## Intro

The articles **Flexible Bayesian Additive Joint Models with an Application to Type 1 Diabetes
Research** [@bamlss:Koehler+Umlauf+Greven:2016] and **Nonlinear Association Structures in Flexible
Bayesian Additive Joint Models** [@bamlss:Koehler+Umlauf+Greven:2018] present a flexible
framework for estimating joint models for longitudinal and survival data using MCMC.
This vignette provides some example code how to set up joint models using _bamlss_.

This vignette explains how to estimate joint models with the following hazard rate model:
$$
h_{i}(t) = \exp\left[\eta_{i}(t)\right] =
   \exp [\eta_{\lambda i}(t) + \eta_{\gamma i} + \eta_{\alpha i}(\eta_{\mu i}(t), t) ]
$$
The longitudinal response $y_{ij}$ at time points $t_{ij}$ is modeled as
$$
y_{ij}=\eta_{\mu i}(t_{ij}) + \varepsilon_{ij}                  
$$   
with independent normal errors $\varepsilon_{ij} \sim N(0, \exp[\eta_{\sigma_i}(t_{ij})]^2)$.

In this model the predictors $\eta_k = \left( \eta^{\lambda}, \eta^{\gamma}, \eta^{\alpha},
\eta^{\mu}, \eta^{\sigma} \right)$ are a composition of
$$
\eta_k=\eta_k\left(\mathbf{x}; \boldsymbol{\beta}_k\right) = f_k(\mathbf{x}) =
  f_{1k}(\mathbf{x}_{1k})+\ldots+f_{J_kk}(\mathbf{x}_{J_kk}),
$$
where functions $f_{jk}( \cdot )$, $j = 1, \ldots, J_k$ are unspecified functions of generic
covariates $\mathbf{x}$, e.g., that are modeled by splines. There are two types of parameters
in this model, time-varying ($\lambda$, $\alpha$ and $\mu$) and time-constant
($\gamma$ and $\sigma$) parameters. Note that parameter $\lambda$ includes the baseline hazard
effect.

For more details on model specifications and methodology see
@bamlss:Koehler+Umlauf+Greven:2016 and @bamlss:Koehler+Umlauf+Greven:2018.

## Mayo Clinic Primary Biliary Cirrhosis Data

This example is taken from the _JMbayes_ package [@bamlss:JMbayes:2016]. The serum bilirubin level
is one of the most important body markers associated with the PBC liver disease. Here, we try to
investigate how the mean bilirubin level influences survival. Therefore, we decompose the
time-varying effect in parameter $\mu$ into a main effect and a functional random intercept, to
account for individual deviations.

The data can be loaded with
```{r}
data("pbc2", package = "JMbayes")
head(pbc2)
```
and contains several potentially relevant covariates explaining the disease. Note that for the
survival part, the time variable is `years` and for the longitudinal part `year` in this
data set. To set up the joint model we need to specify formulas for each parameter
($\lambda$, $\gamma$, $\mu$, $\sigma$, $\alpha$) of the model.
```{r}
## Set up the model formula including
## functional random intercepts using ti().
f <- list(
  Surv2(years, status2, obs = log(serBilir)) ~ s(years,k=20),
  gamma ~ s(age,k=20) + drug + sex,
  mu ~ ti(id,bs="re") + 
       ti(year,k=20) + 
       ti(id,year,bs=c("re","cr"),k=c(nlevels(pbc2$id),8)),
  sigma ~ 1,
  alpha ~ s(years,k=20),
  dalpha ~ -1
)
```
The model is estimated by first running a backfitting algorithm, to find appropriate
starting values, afterwards the MCMC sampler is started.
```{r, eval=FALSE}
## Set the seed for reproducibility.
set.seed(123)

## Estimate model.
b <- bamlss(f, data = pbc2, family = "jm",
  timevar = "year", idvar = "id")
```
```{r, echo=FALSE, results="hide", message=FALSE}
if(!file.exists("figures/JointModel.rda")) {
  set.seed(123)
  b <- bamlss(f, data = pbc2, family = "jm",
    timevar = "year", idvar = "id")
  save(b, file = "figures/JointModel.rda")
} else {
  load("figures/JointModel.rda")
}
```
The model summary gives
```{r}
summary(b)
```
Here, the estimated parameters from the backfitting step are quite different from the posterior
mean calculated from the MCMC samples. This might indicate that the default number of iterations,
`n.iter = 1200`, is not enough to obtain convergence of the MCMC chains. To check this issue
we look at some traceplots.
```{r, eval=FALSE}
plot(b, which = "samples")
```
```{r, fig.width = 9, fig.height = 5, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE}
bsp <- b
bsp$samples <- bsp$samples[[1]][, c("gamma.p.model.matrix.(Intercept)", "mu.p.model.matrix.(Intercept)")]
plot(bsp, which = "samples")
```
For convenience, we only show the traceplots of the intercept for parameter $\gamma$ and $\mu$.
The traceplot for parameter $\mu$ clearly shows an upwards trend of the MCMC chain. To overcome
this issue, we estimate the model again and increase the burnin-in phase and the thinning parameter.
```{r, echo=FALSE, results="hide", message=FALSE}
if(!file.exists("figures/JointModel.rda")) {
  set.seed(321)
  m <- bamlss(f, data = pbc2, family = "jm",
    timevar = "year", idvar = "id",
    n.iter = 12000, burnin = 2000, thin = 10)
  save(m, file = "figures/JointModel2.rda")
} else {
  load("figures/JointModel2.rda")
}
```
```{r, eval=FALSE}
set.seed(321)

## Run model
m <- bamlss(f, data = pbc2, family = "jm",
  timevar = "year", idvar = "id", optimizer = FALSE,
  jm.start = coef(b), n.iter = 12000, burnin = 2000, thin = 10)
```
Note, executing the code will take some time because of the large number of parameters for the
functional random intercept terms ($> 2500$). Another solution to speed up estimation is
to use multiple cores for sampling. This can be done using the `bamlss()` wrapper by setting, e.g.,
`cores = 4`. Then, internally the `mclapply()` function of the _parallel_ [@bamlss:R] package is
called. Now, the samples from model `m` look much better.
```{r, eval=FALSE}
plot(m, which = "samples", lag = 100)
```
```{r, fig.width = 9, fig.height = 5, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE}
bsp <- m
bsp$samples <- bsp$samples[[1]][, c("gamma.p.model.matrix.(Intercept)", "mu.p.model.matrix.(Intercept)")]
plot(bsp, which = "samples", lag = 100)
```
However, there is still some autocorrelation for parameter $\mu$, which could be further reduced by,
e.g., increasing the thinning parameter.

The estimated effects can be plotted instantly, e.g., for the time-varying parameters $\lambda$
and $\alpha$:
```{r, eval=FALSE}
plot(m, model = c("lambda", "alpha"))
```
```{r, fig.width = 7, fig.height = 3, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE, out.width="80%"}
par(mar = c(4.1, 4.1, 1.1, 1.1), mfrow = c(1, 2))
plot(m, model = c("lambda", "alpha"))
```
The left plot is the estimated baseline hazard effect, the right plot the estimated association
parameter. Both effects seem to be quite linear, however, according the 95% credible intervals the
effect for parameter `alpha` is not significant.

The `age` effect for the time-constant parameter $\gamma$ of the risk model is plotted with:
```{r, eval=FALSE}
plot(m, model = "gamma")
```
```{r, fig.width = 5, fig.height = 4, fig.align = "center", dev = "png", results='hide', message=FALSE, echo=FALSE, out.width="50%"}
par(mar = c(4.1, 4.1, 1.1, 1.1))
plot(m, model = "gamma")
```
Again, the effect of age on the logarithmic risk scale is basically linear and increasing. The
estimated effects for the bodymarker model in the $\mu$ parameter are plotted in similar fashion:
```{r, eval=FALSE}
plot(m, model = "mu")
```
```{r, fig.width = 8, fig.height = 2.4, fig.align = "center", dev = "png", results='hide', message=FALSE, echo=FALSE, out.width="100%"}
par(mfrow = c(1, 3), mar = c(4.1, 4.1, 0.1, 1.1))
plot(m, model = "mu", term = c("ti(id)", "ti(id,year)", "ti(year)"))
```
The left plot shows the estimated random effect for each individual, the middle plot the
estimated functional random intercepts and the right plot the corresponding main effect of
`year`, which is linearly increasing.

Another feature that is implemented is the plot of estimated survival probabilities for
single individuals.
```{r, eval=FALSE}
jm_survplot(m, id = 21)
```
```{r, fig.width = 5, fig.height = 6, fig.align = "center", dev = "png", results='hide', message=FALSE, echo=FALSE, out.width="50%"}
jm_survplot(m, id = 21)
```
The upper plot shows the predicted survival probability up to 14 years of individual 21. The lower
plot is the predicted effect of time in the longitudinal model $\mu$. Here, the dots represent
the observed bilirubin level of this individual.

## References

