---
title: "LASSO-Type Penalization in the Framework of Generalized Additive Models
  for Location, Scale and Shape"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{LASSO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{distributional regression, LASSO, fused LASSO}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
set.seed(123)

if(!file.exists("figures/lasso.rda")) {
  file_url <- "http://www.uni-goettingen.de/de/document/download/64c29c1b1fccb142cfa8f29a942a9e05.raw/rent99.raw"
  rent99 <- read.table(file_url, header = TRUE)
  rent99$yearc <- ordered(floor(rent99$yearc/10)*10)
  rent99$area <- cut(rent99$area, breaks = seq(20, 160, by = 20),
    include.lowest = TRUE, right = FALSE, ordered = TRUE)
  rent99$location <- as.factor(rent99$location)
  f <- list(
    rentsqm ~ la(area,fuse=2) + la(yearc,fuse=2) + la(location,fuse=1),
    sigma ~ la(area,fuse=2) + la(yearc,fuse=2) + la(location,fuse=1)
  )
  rentmodel <- bamlss(f, data = rent99, optimizer = opt_lasso, sampler = FALSE,
    criterion = "BIC", nlambda = 100, multiple = TRUE, upper = exp(6), lower = exp(0))

  rentmodel_s <- bamlss(f, data = rent99, optimizer = opt_bfit, sampler = FALSE,
    criterion = "BIC", nu = 1, start = coef(rentmodel, mstop = lasso_stop(rentmodel)))

  save(rentmodel, rentmodel_s, file = "figures/lasso.rda")
} else {
  load("figures/lasso.rda")
}
```

## Intro

The article **LASSO-type penalization in the framework of generalized additive models for
location, scale and shape** [@bamlss:Groll+Hambuckers+Kneib+Umlauf:2019] presents
a regularization approach for high-dimensional data set-ups in the framework of GAMLSS.
It is designed for linear covariate effects and is based on L$_{1}$-type penalties. The 
following three penalization options are provided:

* the conventional least absolute shrinkage and selection operator (LASSO) for metric
covariates,
* the group LASSO,
* and fused LASSO for categorical predictors.

This vignette provides some example code how to set up LASSO-type penalized models using _bamlss_.

## Munich rental guide data

We follow the example in the article [@bamlss:Groll+Hambuckers+Kneib+Umlauf:2019] and use
Munich rental guide data to illustrate the modeling approach. Unfortunately, we cannot use the
original data shown in the paper because of copywright reasons. However, an older version of the
data is available in the supplementary materials of the Regression book
[@bamlss:Fahrmeir+Kneib+Lang+Marx:2013]. The data can be loaded into R with
```{r}
file_url <- "http://www.bamlss.org/misc/rent99.raw"
rent99 <- read.table(file_url, header = TRUE)
```
The response is the monthly rent per square meter in Euro (`rentsqm`). In this example
we use the year of construction (`yearc`), area (`area`) and the location classification
(`location', derived from expert knowledge) of the apartment as covariates. Following the example
[@bamlss:Groll+Hambuckers+Kneib+Umlauf:2019], the two continuous covariates are categorized 
```{r}
## Create factors of continuous covariates.
rent99$yearc <- ordered(floor(rent99$yearc / 10) * 10)
rent99$area <- cut(rent99$area, breaks = seq(20, 160, by = 20),
  include.lowest = TRUE, right = FALSE, ordered = TRUE)

## Also transform to factor.
rent99$location <- as.factor(rent99$location)
```

We fit a Gaussian GAMLSS and use for both distribution parameters, i.e.\ $\mu$ and $\sigma$, using
the following code
```{r, eval=FALSE}
## Model formula.
f <- list(
  rentsqm ~ la(area,fuse=2) + la(yearc,fuse=2) + la(location,fuse=1),
  sigma ~ la(area,fuse=2) + la(yearc,fuse=2) + la(location,fuse=1)
)

## LASSO model based on lambda grid,
## use special opt_lasso() estimation engine.
rentmodel <- bamlss(f, data = rent99, optimizer = opt_lasso, sampler = FALSE,
  criterion = "BIC", nlambda = 100, multiple = TRUE, upper = exp(6), lower = exp(0))
```
where arguments `upper` and `lower` define the limits of the LASSO tuning parameters. Estimation
in this case is based on the BIC using a two dimensional grid search with $200 \times 200$ iterations.
Note that `fuse=1` is used for nominal fusion and `fuse=2` for ordinal fusion within the special
`la()` model term constructor function.

The resulting marginal BIC curves can be visualized with
```{r, eval=FALSE}
pathplot(rentmodel, which = "criterion")
```
```{r, fig.width = 8, fig.height = 4, fig.align = "center", echo = FALSE, dev = "png", echo=FALSE}
par(mfrow = c(1, 2))
pathplot(rentmodel, which = "criterion", spar = FALSE)
```
In this example both tuning parameters have the same value, also indicated by the gray dashed
vertical lines. The coefficient paths w.r.t.\ the tuning parameters can be plotted, e.g.,
for parameter $\mu$, with
```{r, eval=FALSE}
pathplot(rentmodel, which = "parameters", model = "mu")
```
```{r, fig.width = 5, fig.height = 5, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE}
par(mfrow = c(1, 2))
pathplot(rentmodel, which = "parameters", model = "mu")
```
The legend on the right hand side of the plot is not necessarily very nice (too long). One option
to select specific paths for plotting is to set the `name` argument in `pathplot()`, e.g.,
plotting the coefficient paths for the year of construction for parameter $\mu$ can be done by
```{r, fig.width = 5, fig.height = 5, fig.align = "center", dev = "png", results = 'hide'}
pathplot(rentmodel, which = "parameters", model = "mu",
  name = "mu.s.la(yearc).yearc")
```
The plotting function also replaces the names of the coefficients such that the labeling
in the plot is easy to read. The plot shows that very old apartments are all shrinked out
of the model, whereas apartments from 1950 and younger enter the model and categories are
not fused.

Similarly, the effect on parameter $\sigma$ for the `area` of the apartment can be plotted by
```{r, fig.width = 5, fig.height = 5, fig.align = "center", dev = "png", results = 'hide'}
pathplot(rentmodel, which = "parameters", model = "sigma",
  name = "sigma.s.la(area).area")
```
The plot shows that almost all categories of covariate `area` are fused using the BIC for tuning
parameter selection and uncertainty seems to be lowest for larger apartments.

## Single tuning parameters.

In _bamlss_, it is also possible to estimate single tuning parameters for each LASSO term.
The main advantage of the single tuning parameter selection is that in principle classical
LASSO and fused LASSO terms can be mixed, which is not the case if only single tuning parameters,
one for each distributional parameter (e.g., $\mu$ and $\sigma$ using the Gaussian distribution),
is used.

E.g., using the Munich rent data, we could estimate the model with the default backfitting
model fitting engine `opt_bfit()`
```{r, eval=FALSE}
rentmodel_s <- bamlss(f, data = rent99, optimizer = opt_bfit, sampler = FALSE,
  criterion = "BIC", nu = 1, start = coef(rentmodel, mstop = lasso_stop(rentmodel)))
```
using the estimated parameters of the first model as starting values (`mstop` is the optimum
stopping iteration of the `opt_lasso()` optimizer). Note that we also optimize the step length of
the backfitting algorithm by setting `nu = 1`. To compare both models by BIC, the `lasso_stop()`
functions returns the optimum iteration of the `opt_lasso()` optimizer
```{r}
lasso_stop(rentmodel)
```
along the estimated degrees of freedom and the BIC. Similarly, the BIC of the second model can
be extracted with
```{r}
BIC(rentmodel_s)
```
which is slightly a bit smaller than the first model suggesting that the grid search method
is most probably enough in this case.

Using the `lasso_stop()` function, we can also extract the corresponding predictions.
```{r, eval=FALSE}
predict(rentmodel, mstop = lasso_stop(rentmodel))
```
Note that the default in `coef()` and `predict()` would be to use the last LASSO iteration
otherwise.

## References

