---
title: "Bayesian Multinomial Regression"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{LASSO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{distributional regression, multinomial model}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
set.seed(123)
data("marital.nz", package = "VGAM")

if(!file.exists("figures/multinomial.rda")) {
  f <- list(
    mstatus ~ s(age),
            ~ s(age),
            ~ s(age)
  )
  set.seed(123)
  b <- bamlss(f, family = "multinomial", data = marital.nz, reference = "Married/Partnered")
  save(b, file = "figures/multinomial.rda")
} else {
  load("figures/multinomial.rda")
}
```

This vignette is based on @bamlss:Yee:2009. The data is about the marital status of white male
in New Zealand in the early 1990s. The aim of this analysis is to explore how the marital status
varies with age. The data can be loaded with
```{r}
data("marital.nz", package = "VGAM")
head(marital.nz)
levels(marital.nz$mstatus)
```
The response `mstatus` has 4 levels. We use a multinomial logit model to estimate the age effect,
therefore, one category needs to be specified as a reference category. The model can be
estimated with
```{r, eval=FALSE}
## Model formula, each category may
## have different model terms.
f <- list(
  mstatus ~ s(age),
          ~ s(age),
          ~ s(age)
)

## Set the seed for reproducibility.
set.seed(123)

## Estimate.
b <- bamlss(f, family = "multinomial", data = marital.nz,
  reference = "Married/Partnered")
```
The model summary gives
```{r}
summary(b)
```
and suggests reasonable acceptance rates. However, for a final model run it is recommend to
increase the number of iteration, the burn-in phase as well as the thinning parameter of the
sampling engine function `GMCMC()`. The estimated effects can be plotted with
```{r, fig.width = 8, fig.height = 3, fig.align = "center", dev = "png", out.width="100%"}
par(mfrow = c(1, 3), mar = c(4.1, 4.1, 0.1, 1.1))
plot(b)
```
To calculate estimated age dependent probabilities for each category, we use the `predict.bamlss()`
method.
```{r}
## Create a data frame for prediction.
nd <- data.frame("age" = seq(20, 90, length = 100))

## Predict for the three levels.
p <- predict(b, newdata = nd, type = "parameter")
```
Now, note that the specification of the predictors in the `multinomial_bamlss()` family
is based on logarithmic link, therefore, to compute the probabilities we run the
following code:
```{r}
probs <- list()
for(j in names(p))
  probs[[j]] <- p[[j]] / (1 + rowSums(do.call("cbind", p)))
probs <- as.data.frame(probs)
probs[["MarriedPartnered"]] <- 1 - rowSums(probs)
```
The estimated probabilities can then be visualized with:
```{r, fig.width = 6, fig.height = 5, fig.align = "center", dev = "png", out.width="60%"}
par(mar = c(4.1, 4.1, 0.1, 0.1))
plot2d(probs ~ age, data = nd, col.lines = rainbow_hcl(4),
  lwd = 2, scheme = 1, ylab = "Fitted probabilities", ylim = c(0, 1))
legend("center", names(probs), lty = 1, lwd = 2, col = rainbow_hcl(4))
```

## References

