---
title: "Munich Rent Model"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{Model Terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{distributional regression, Munich rent data}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
set.seed(123)

if(!file.exists("figures/rent99.rda")) {
  rent99 <- read.table("http://www.uni-goettingen.de/de/document/download/64c29c1b1fccb142cfa8f29a942a9e05.raw/rent99.raw",
    header = TRUE)
  save(rent99, file = "figures/rent99.rda")
} else {
  load("figures/rent99.rda")
}

if(!file.exists("figures/rentmodels.rda")) {
  f1 <- rent ~ area + yearc

  b1 <- bamlss(rent ~ area + yearc, data = rent99, family = "gaussian",
    n.iter = 12000, burnin = 2000, thin = 10)

  f2 <- list(
    rent ~ s(area,k=20) + s(yearc,k=20),
    sigma ~ s(area,k=20) + s(yearc,k=20)
  )

  b2 <- bamlss(f2, data = rent99, family = "gaussian",
    n.iter = 12000, burnin = 2000, thin = 10)

  save(b1, b2, file = "figures/rentmodels.rda")
} else {
  load("figures/rentmodels.rda")
}
```

In this example we analyze rent index data from Munich, Germany. The data is provided in the
supplemental materials of the Regression book [@bamlss:Fahrmeir+Kneib+Lang+Marx:2013] and can be
loaded into R with
```{r}
file_url <- "http://www.uni-goettingen.de/de/document/download/64c29c1b1fccb142cfa8f29a942a9e05.raw/rent99.raw"
rent99 <- read.table(file_url, header = TRUE)
```

The aim of this analysis is to find a model explaining the rent or rent per square meter of an
apartment. We first assume a linear relationship with homoscedastic Gaussian errors:

$$
\texttt{rent} = \mu = \beta_0 + \beta_1 \cdot \texttt{area} + \beta_2 \cdot \texttt{yearc},
$$
i.e., $\texttt{rent} \sim N(\mu, \sigma^2)$. The model can be estimated with: 

```{r, eval=FALSE}
## Model formula.
f1 <- rent ~ area + yearc

## Estimate model.
b1 <- bamlss(f1, data = rent99, family = "gaussian",
  n.iter = 12000, burnin = 2000, thin = 10)
```
This first starts a backfitting algorithm to find posterior mode estimates and afterwards the MCMC
simulation is started using the parameters from the backfitting step as starting values.
Note that we use 12000 iterations with a burnin of 2000 and only keep every 10th sample of
the MCMC simulation. The model summary gives
```{r}
summary(b1)
```
However, the assumption of homoscedastic errors is not appropriate in this case.
```{r, eval=FALSE}
plot(rent ~ area, data = rent99, pch = 16, col = rgb(0.3, 0.3, 0.3, alpha = 0.3))
```
```{r, fig.width = 5, fig.height = 4, fig.align = "center", dev = "png", results='hide', message=FALSE, echo=FALSE, out.width="50%"}
par(mar = c(4.1, 4.1, 1.1, 1.1))
plot(rent ~ area, data = rent99, pch = 16, col = rgb(0.3, 0.3, 0.3, alpha = 0.3))
```
As can be seen in this scatter plot, the data is heteroscedastic and the exact model might not even
be linear. This is also indicated in the quantile residual plots.
```{r, eval=FALSE}
plot(b1, which = c("hist-resid", "qq-resid"))
```
```{r, fig.width = 7.5, fig.height = 4, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE, out.width="80%"}
par(mfrow = c(1, 2))
plot(b1, which = c("hist-resid", "qq-resid"), spar = FALSE)
```
The residuals seem to be a bit skew and for the lower and higher values of `rent`, the model
fit is not satisfactorily according the Q-Q plot.

A non-parametric model can be useful here. Therefore, we estimate a more flexible model
with
$$
\mu = \beta_0 + f_1(\texttt{area}) + f_2(\texttt{yearc})
$$
and
$$
\sigma = \exp\{\alpha_0 + g_1(\texttt{area}) + g_2(\texttt{yearc})\}
$$
where functions $f_1( \cdot ), f_2( \cdot ), g_1( \cdot ), g_2( \cdot )$ are modeled using regression
splines. The exponential function assures that the we estimate a positive standard deviation. This way,
we allow for covariates and loose the hard assumptions of a constant standard deviation. 

The model is estimated with: 
```{r, eval=FALSE}
## List of formulae to estimate
## heteroskedastic Gaussian model.
f2 <- list(
  rent ~ s(area,k=20) + s(yearc,k=20),
  sigma ~ s(area,k=20) + s(yearc,k=20)
)

b2 <- bamlss(f2, data = rent99, family = "gaussian",
  n.iter = 12000, burnin = 2000, thin = 10)
```
```{r}
summary(b2)
```
The summary statistics indicate that the more flexible model is more appropriate since the
DIC of model `b2` is much lower than the DIC of model `b1`.

The estimated effects are visualized with
```{r, eval=FALSE}
plot(b2, ask = FALSE)
```
```{r, fig.width = 7.2, fig.height = 5.8, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE, out.width="80%"}
par(mar = c(4.1, 4.1, 1.1, 1.1), mfrow = c(2, 2))
plot(b2, ask = FALSE)
```
By looking at the effect plots, we can conclude that in the case of the variable `area` an
assumption of linear dependency might be justified, where in the case of `yearc` this is not so clear.
Moreover, our estimated `sigma` is for neither of the variables constant.
```{r, eval=FALSE}
plot(b2, which = c("hist-resid", "qq-resid"))
```
```{r, fig.width = 7.5, fig.height = 4, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE, out.width="80%"}
par(mfrow = c(1, 2))
plot(b2, which = c("hist-resid", "qq-resid"), spar = FALSE)
```
The quantile residual plots indicate a quite good model fit already, however, the assumption
of Gaussian errors could be refined as rents can only be positive. Hence, distributions
accounting for this fact should be exploited further, e.g., the Gamma distribution.

