---
title: "Available Model Terms"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{Model Terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{distributional regression, model terms}
  %\VignettePackage{bamlss}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
set.seed(123)
```

## Overview

The _bamlss_ package heavily builds upon the R package _mgcv_ [@bamlss:Wood:2018] infrastructures,
i.e., all model terms that are provided by _mgcv_ can also be used in _bamlss_. As a consequence,
it is also possible to write user defined model terms using the generic `smooth.construct()`method
of _mgcv_. Moreover, users can in principle even extent this, writing model terms beyond _mgcv_
framework. However, this is a bit more technical and assumes that the user has a good basic
knowledge of the internal structure of the package and is only very briefly discussed in this
article.

## Specification

To give an overview of users not familiar with _mgcv_, the following table lists some model
specifications using R's formula syntax:

| Description | Formula |
|:---|:---|
| Linear effects $\mathbf{X}\boldsymbol{\beta}$ | `~ x1 + x2 + x3` |
| Nonlinear effects of continuous covariates $f(\mathbf{x}) = f(x_1)$ | `~ s(x1)` |
| Two-dimensional surfaces $f(\mathbf{x}) = f(x_1, x_2)$ | `~ s(x1,x2)`, `~ te(x1,x2)` or `~ ti(x1,x2)` (higher dimensional terms possible) |
| Spatially correlated effects $f(\mathbf{x}) = f_{spat}(x_s)$ | `~ s(xs,bs="mrf",xt=list(penalty=K))`, where `xs` is a factor indicating the discrete regional information and `K` is a supplied penalty matrix. Other options within the `xt` argument are possible, please see the documentation of `smooth.construct.mrf.smooth.spec()` |
| Varying coefficients $f(\mathbf{x}) = x_1f(x_2)$ | `~ s(x2,by=x1)` |
| Spatially varying effects $f(\mathbf{x}) = x_1f_{spat}(x_s)$ or $f(\mathbf{x}) = x_1f(x_2, x_3)$ | `~ s(xs,bs="mrf",xt=list(penalty=K),by=x1)`, `~ s(x2,x3,by=x1)` or `~ te(x2,x3,by=x1)` |
| Random intercepts with cluster index $c$: $f(\mathbf{x}) = \beta_c$ | `~ s(id,bs="re")`, where `id` is a factor of cluster indices |
| Random slopes with cluster index $c$: $f(\mathbf{x}) = x_1\beta_c$ | `~ s(id,x1,bs="re")`, as above with continuous covariate `x1` |

Note that each model term constructor has an argument called `bs`, which specifies the type
of basis that should be used for modeling.
For more details on the smooth constructors, please visit the _mgcv_ manual pages of `s()`, `te()` and `ti()`.

## Basics

To better introduce the standard types of model terms we use simulated data that is part of the
_bamlss_ package. The data contains of a couple of different response types and is usually
used for internal tests, only.
```{r}
d <- GAMart()
print(head(d))
```
The first column holding variable `num` is a Gaussian response, all subsequent columns up to
variable `cens` are basically transformations of variable `num`. Variables `x1` to `lat` are
covariates, which we use for illustrating how to set up the different model terms in the
following.

Let's start with a simple example using three covariates for a Gaussian model with
possible nonlinear influence on the response distribution. The model formula can be specified by
```{r}
f <- list(
  num ~ s(x1) + s(x2) + s(x3),
  sigma ~ s(x1) + s(x2) + s(x3)
)
```
Note that the default basis function type in _mgcv_ are thin plate regression splines
[@bamlss:Wood:2003] using `k = 10` basis functions per default. Setting `k = 10` is reasonable in
most cases, but this may not be enough for very complex functional forms.

The type of basis in _mgcv_ is selected with the `bs` argument. There are a number of different
basis types readily available, from P-splines [@bamlss:Eilers+Marx:1996] to Markov random fields,
this is documented on the _mgcv_ manual page `?smooth.terms`.
Setting up the smooth terms is relatively straightforward when already known one. For example,
if one is interested in specifying interaction terms, e.g., with main effects and interaction using
two covariates, we could do the following using tensor product splines
```{r}
f <- list(
  num ~ ti(x1) + ti(x2) + ti(x1,x2),
  sigma ~ s(x1) + s(x2) + s(x3)
)
```
Here, we use a functional decomposition of the effects of covariates `x1` and `x2` in the
first formula. Note again that the number of basis function within `ti()` is again small, `k = 5`.
The reason is, that constructing tensor product model terms can easily end up with thousands of
parameters, i.e., these type of model terms need to be specified with care.

## Markov Random Fields

Markov random fields (MRF) can be used to model spatially correlated effects using discrete
locational observations. However, setting up MRFs can be a bit tricky, so I would like to illustrate
the process with a short example. In the simulated data from above, we artificially sample from
spatial locations, see the plot of the coordinates
```{r, eval=FALSE}
plot(lat ~ lon, data = d)
```
```{r, fig.width = 4.5, fig.height = 4, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE}
par(mar = c(4.1, 4.1, 0.1, 0.1))
plot(lat ~ lon, data = d)
```
Let's suppose we only have the coordinates of the locations. Sure, we could directly model the
spatial effect using some 2D smoother, e.g., with thin plate regression splines. In most situations
spatial information is available as region identifier, but for illustrating the full path, it is
a good exercise to start from exact locations. Now, either we have a map available

## References

