---
title: "Spatial location-scale model"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: bamlss.bib
nocite: '@bamlss:Umlauf+bamlss:2018'
vignette: >
  %\VignetteIndexEntry{Model Terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{bamlss}
  %\VignetteKeywords{distributional regression, GLM, big data}
  %\VignettePackage{bamlss}
---


```{r preliminaries, echo=FALSE, message=FALSE}
library("bamlss")
set.seed(123)
```
This example is taken from the _R2BayesX_ package [@bamlss:Umlauf+Adler+Kneib+Lang+Zeileis:2014] and
is about undernutrition of new born children in Zambia. The data is loaded with
```{r}
data("ZambiaNutrition", package = "R2BayesX")
head(ZambiaNutrition)
```
Here, the primary interest is to model the dependence of `stunting` of newborn children, with an
age ranging from 0 to 5 years, on covariates such as the body mass index of the mother, the age of
the child and others. Moreover, we apply a full distributional regression model with
$$
\texttt{stunting} \sim \mathcal{N}(\mu = \eta_{\mu}, \log(\sigma) = \eta_{\sigma})),
$$
where the predictors $\eta_{\mu}$ and $\eta_{\sigma}$ are specified by the following formula
```{r}
f <- list(
  stunting ~ memployment + urban + gender + meducation +
    s(mbmi) + s(agechild) + s(district, bs = "mrf", xt = list("penalty" = K)) +
    s(district, bs = "re"),
  sigma  ~ memployment + urban + gender + meducation +
    s(mbmi) + s(agechild) + s(district, bs = "mrf", xt = list("penalty" = K)) +
    s(district, bs = "re")
)
```
Note that for setting up the Markov random field smooth term a penalty matrix `K` needs to be
provided. The penalty matrix forces penalization for neighboring regions of the districts in Zambia.
To compute the `K` matrix, we need the spatial information about the regions in Zambia, which
is shipped as a `"bnd"` object in the _R2BayesX_ package and can be loaded with
```{r}
data("ZambiaBnd", package = "R2BayesX")
```
The `K` matrix can then be computed using function `neighbormatrix()`
```{r}
K <- neighbormatrix(ZambiaBnd)
head(K)

## Also need to transform to factor for
## setting up the MRF smooth.
ZambiaNutrition$district <- as.factor(ZambiaNutrition$district)

## Now note that not all regions are observed,
## therefore we need to remove those regions
## from the penalty matrix
rn <- rownames(K)
lv <- levels(ZambiaNutrition$district)
i <- rn %in% lv
K <- K[i, i]
```
Then, the model can be estimated with
```{r, eval=FALSE}
set.seed(321)
b <- bamlss(f, data = ZambiaNutrition, family = "gaussian")
```
and the estimated univariate effects are plotted with
```{r, eval=FALSE}
plot(b)
```
```{r, fig.width = 7.2, fig.height = 5.8, fig.align = "center", echo = FALSE, dev = "png", results = 'hide', message=FALSE, out.width="80%"}
par(mar = c(4.1, 4.1, 1.1, 1.1), mfrow = c(2, 2))
if(!file.exists("figures/ZambiaModel.rda")) {
  set.seed(321)
  b <- bamlss(f, data = ZambiaNutrition, family = "gaussian")
  save(b, file = "figures/ZambiaModel.rda")
} else {
  load("figures/ZambiaModel.rda")
}
plot(b, ask = FALSE)
```
The plot indicates that only the effect of variable `mbmi` on the standard deviation is not
significant according the 95% credible intervals and basically follows the zero horizontal line.

To visualize the structured and unstructured spatial effects we predict using the district
information
```{r}
## First, note that we have the structured id = 'mrf1' and unstructured
## spatial effect id = 're2', also indicated in the model summary
summary(b)

## Now, to predict the spatial effects we set up new data.
nd <- data.frame("district" = unique(ZambiaNutrition$district))

## Predict for the structured spatial effects.
p_str <- predict(b, newdata = nd, term = "s(district,id='mrf1')", intercept = FALSE)

## And the unstructured spatial effect.
p_unstr <- predict(b, newdata = nd, term = "s(district,id='re2')", intercept = FALSE)
```
Now, to visualize the effects we plot all maps using the same range
for the color legends
```{r}
r_mu <- range(c(p_str$mu, p_unstr$mu))
r_mu <- c(-1 * max(abs(r_mu)), max(abs(r_mu)))

r_sigma <- range(c(p_str$sigma, p_unstr$sigma))
r_sigma <- c(-1 * max(abs(r_sigma)), max(abs(r_sigma)))
```
and plot the effects using a diverging color legend.
```{r, eval=FALSE}
## MRF smooth effect.
plotmap(ZambiaBnd, x = p_str$mu, id = nd$district, color = diverge_hcl, range = r_mu,
  main = expression(mu), shift = 0.1, title = "MRF", mdensity = 20)
plotmap(ZambiaBnd, x = p_str$sigma, id = nd$district, color = diverge_hcl, range = r_sigma,
  main = expression(sigma), shift = 0.1, title = "MRF", mdensity = 20)

## Random effects.
plotmap(ZambiaBnd, x = p_unstr$mu, id = nd$district, color = diverge_hcl, range = r_mu,
  shift = 0.1, title = "Random effect", mdensity = 20)
plotmap(ZambiaBnd, x = p_unstr$sigma, id = nd$district, color = diverge_hcl, range = r_sigma,
  shift = 0.1, title = "Random effect", mdensity = 20)
```
```{r, fig.width = 7, fig.height = 7, fig.align = "center", dev = "png", results='hide', message=FALSE, echo=FALSE, out.width="60%"}
par(mfrow = c(2, 2), mar = c(0, 0, 4.1, 0))
plotmap(ZambiaBnd, x = p_str$mu, id = nd$district,
  color = diverge_hcl, range = r_mu, shift = 0.1, title = "MRF",
  mdensity = 20)
mtext(expression(mu), side = 3, line = 1, cex = 2)
plotmap(ZambiaBnd, x = p_str$sigma, id = nd$district,
  color = diverge_hcl, range = r_sigma, shift = 0.1, title = "MRF",
  mdensity = 20)
mtext(expression(sigma), side = 3, line = 1, cex = 2)

plotmap(ZambiaBnd, x = p_unstr$mu, id = nd$district,
  color = diverge_hcl, range = r_mu, mdensity = 20,
  shift = 0.1, title = "Random effect")
plotmap(ZambiaBnd, x = p_unstr$sigma, id = nd$district,
  color = diverge_hcl, range = r_sigma, mdensity = 20,
  shift = 0.1, title = "Random effect")
```
The maps clearly show that the unstructured spatial effect seems to very small, if existent
at all when looking the 95% credible intervals:
```{r}
## Again predict, but now additionally compute 95% credible intervals
## using function c95().
p_unstr <- predict(b, newdata = nd, term = "s(district,id='re2')", intercept = FALSE, FUN = c95)

## Test if all effects contain zero, i.e., are not significant
## according the 95% credible intervals.
all(p_unstr$mu[["2.5%"]] < 0 & p_unstr$mu[["97.5%"]] > 0)
all(p_unstr$sigma[["2.5%"]] < 0 & p_unstr$sigma[["97.5%"]] > 0)
```

## References

